\rm Weak compactness of probability measures

### 2018-07-04 manuscript-3.tex
Adding Klenke as reference.

In probability theory it is common to consider the \emph{weak topology} on the Borel probability measures \(\mathcal P(\mathcal A)\) over some topological space \(X\) with Borel algebra \(\mathcal A\), which is the initial topology of the evaluations
\[f\mapsto \mu(f) = \int f\mathrm{d}\mu\quad \text{for } f\in\mathcal C_b( X).\]
%We will now try to address the 
Note that this corresponds exactly to the relative topology on the probability measures \(\mathcal P(\mathcal A)\subseteq \mathcal C_b(X)^\ast\)  of the weak-\(\ast\) topology, where \(\mathcal P(\mathcal A)\) is embedded canonically.
If \(X\) is a complete metric space, then it is well known that the convergence that arrises from this topology, which is called the \emph{weak convergence}, can be metrised by the dual bounded Lipschitz metric or the Lévy-Prokhorov metric.
%\[d_{L}(\mu,\nu) \coloneqq \sup \Big\{ \mu(f) - \nu(f) \mid f\in\mathcal C_b(X) \text{ is Lipschitz with constant at most } 1 \Big\}.  \]
However this result is frequently stated as the fact that the weak topology is metrised with by this metric. Against the background of the metrisability results of the weak topologies in Banach spaces this seems very surprising and in fact it turns out to be false. More precisely the weak topology is metrisable if and only if the space \(X\) is compact, which is the case if and only if \(\mathcal C_b(X)\) is separable. This agrees with the metrisability of the weak-\(\ast\) topology.
To prove this statement, we will need the following ingredrients.
\begin{enumerate}
\item The space of probability measures over a topological space is compact in the weak topology.
\item If a famility \(\left\{ \mu_i\right\}_{i\in I}\) of probability measures is relatively sequentially compact in the weak topology, then it is \emph{tight}, i.e. for every \(\varepsilon>0\) there is a compact set \(K\subseteq X\) such that \(\mu_i(K)>1-\varepsilon\) for all \(i\in I\).
\item If \(X\) is a compact metric space, then \(\mathcal C_b(X)\) is separable. Note that the converse statement is also true, but we will not need it here.% if and only if \(X\) is compact.
\end{enumerate}
Assuming we have all those statements, then we can conclude as follows. If \(X\) is compact, then \(\mathcal C_b(X)\) is seperable and thus the weak-\(\ast\) topology on \(\mathcal C_b(X)^\ast\) is metrisable. In this case also the relative topology, i.e. the weak topology on the probability measures is metrisable. If on the other hand the weak topology is metrisable, then we know because of (\emph{i}) that the probability measures are sequentially compact. In particular the family of Dirac measures \(\left\{ \delta_x\right\}_{x\in X}\) is tight and thus there is a compact set \(K\subseteq X\) such that
\[\delta_x(K)>\frac12, \quad \text{i.e. } x\in K.\]
Hence we obtain that \(X=K\) is compact. 
Since (\emph{ii}) is well established in the literature (c.f. Prokhorov’s theorem or Theorem 5.2 in \cite{billingsley2013convergence}) and (\emph{iii}) follows from the Stone-Weierstrass theorem (c.f. \cite{rudin1991functional}), we will focus on the proof of (\emph{i}).

@@

In the last section it was shown that the set of probability measures are compact with respect to the weak topology. Further the motivating question was answered, namely it was the goal to investigate whether the Prokhorov metric metrises the weak topology like frequently stated or only the weak convergence (c.f. \cite{billingsley2013convergence}).% but frequently stated that it would metrise the weak topology itself.
% the question that was the motivation of this work was answered. Namely we wanted  It is well known that the Prokhorov metric metrises the weak convergence (c.f. \cite{billingsley2013convergence}) but frequently stated that it would metrise the weak topology itself.
% This implies that the weak topology of probability measures over a complete and separable metric space is metrisable if and only the space is compact. This answers the question which %was the motivation of the presented work whether the weak topology or only the weak convergence is metrised by the Prokhorov metric (c.f. \cite{billingsley2013convergence}).
In fact the weak topology of probability measures over a complete and separable metric space is metrisable if and only the space is compact and thus only the weak convergence is metrised. However this is no contradiction but shows that the weak topology has no countable neighborhood basis and is therefore not uniquely specified by its notion of convergence\todo{cite}.

>>>>

In probability theory it is common to consider the \emph{weak topology} on the Borel probability measures \(\mathcal P(\mathcal A)\) over some topological space \(X\) with Borel algebra \(\mathcal A\), which is the initial topology of the evaluations
\[f\mapsto \mu(f) = \int f\mathrm{d}\mu\quad \text{for } f\in\mathcal C_b( X).\]
%We will now try to address the 
Note that this corresponds exactly to the relative topology on the probability measures \(\mathcal P(\mathcal A)\subseteq \mathcal C_b(X)^\ast\)  of the weak-\(\ast\) topology, where \(\mathcal P(\mathcal A)\) is embedded canonically.
If \(X\) is a complete metric space, then it is well known that the convergence that arrises from this topology, which is called the \emph{weak convergence}, can be metrised by the dual bounded Lipschitz metric or the Lévy-Prokhorov metric.
%\[d_{L}(\mu,\nu) \coloneqq \sup \Big\{ \mu(f) - \nu(f) \mid f\in\mathcal C_b(X) \text{ is Lipschitz with constant at most } 1 \Big\}.  \]
However this result is sometimes stated as the fact that the weak topology is metrised with by this metric (c.f. \cite{klenke2013probability}). Against the background of the metrisability results of the weak topologies in Banach spaces this seems very surprising and in fact it turns out to be false. More precisely the weak topology is metrisable if and only if the space \(X\) is compact, which is the case if and only if \(\mathcal C_b(X)\) is separable. This agrees with the metrisability of the weak-\(\ast\) topology.
To prove this statement, we will need the following ingredrients.
\begin{enumerate}
\item The space of probability measures over a topological space is compact in the weak topology.
\item If a famility \(\left\{ \mu_i\right\}_{i\in I}\) of probability measures is relatively sequentially compact in the weak topology, then it is \emph{tight}, i.e. for every \(\varepsilon>0\) there is a compact set \(K\subseteq X\) such that \(\mu_i(K)>1-\varepsilon\) for all \(i\in I\).
\item If \(X\) is a compact metric space, then \(\mathcal C_b(X)\) is separable. Note that the converse statement is also true, but we will not need it here.% if and only if \(X\) is compact.
\end{enumerate}
Assuming we have all those statements, then we can conclude as follows. If \(X\) is compact, then \(\mathcal C_b(X)\) is seperable and thus the weak-\(\ast\) topology on \(\mathcal C_b(X)^\ast\) is metrisable. In this case also the relative topology, i.e. the weak topology on the probability measures is metrisable. If on the other hand the weak topology is metrisable, then we know because of (\emph{i}) that the probability measures are sequentially compact. In particular the family of Dirac measures \(\left\{ \delta_x\right\}_{x\in X}\) is tight and thus there is a compact set \(K\subseteq X\) such that
\[\delta_x(K)>\frac12, \quad \text{i.e. } x\in K.\]
Hence we obtain that \(X=K\) is compact. 
Since (\emph{ii}) is well established in the literature (c.f. Prokhorov’s theorem or Theorem 5.2 in \cite{billingsley2013convergence}) and (\emph{iii}) follows from the Stone-Weierstrass theorem (c.f. \cite{rudin1991functional}), we will focus on the proof of (\emph{i}).

@@

In the last section it was shown that the set of probability measures is compact with respect to the weak topology. Further the motivating question was answered, namely it was the goal to investigate whether the Prokhorov metric metrises the weak topology like frequently stated or only the weak convergence (c.f. \cite{billingsley2013convergence}).% but frequently stated that it would metrise the weak topology itself.
% the question that was the motivation of this work was answered. Namely we wanted  It is well known that the Prokhorov metric metrises the weak convergence (c.f. \cite{billingsley2013convergence}) but frequently stated that it would metrise the weak topology itself.
% This implies that the weak topology of probability measures over a complete and separable metric space is metrisable if and only the space is compact. This answers the question which %was the motivation of the presented work whether the weak topology or only the weak convergence is metrised by the Prokhorov metric (c.f. \cite{billingsley2013convergence}).
In fact the weak topology of probability measures over a complete and separable metric space is metrisable if and only the space is compact and thus only the weak convergence is metrised. However this is no contradiction but shows that the weak topology has no countable neighborhood basis and is therefore not uniquely specified by its notion of convergence\todo{cite}.
